Gammapy: A Python package for gamma-ray astronomy

### 2023-03-03 src/ms-review.tex
Fix #95

%% \gammaray astronomy
%% CN: I think we should use the expression Water Cherenkov Detectors and use the
%% abbreviation (WCD) all along the text, later we use "particle samplers"
The \gammaray range of the electromagnetic spectrum provides us insights into the
most energetic processes in the universe such as those accelerating particles in the surroundings of
black holes, and remnants of supernova explosions. As in other
branches of astronomy, \gammarays can be observed by both
satellite as well as ground based instruments.
Ground-based instruments use the Earth's atmosphere as a particle detector.
Very-high-energy (VHE) cosmic \gammarays interact in the atmosphere and
create a large shower of secondary particles that can be observed from the ground.
Ground-based \gammaray astronomy relies on these extensive air showers to detect the
primary \gammaray photons and infer their incident direction and energy.
VHE \gammaray astronomy covers the energy range from fews tens of ${\rm GeV}$ up to the ${\rm PeV}$.
There are two main categories of ground-based instruments:

@@

\textit{Water Cherenkov Detectors (WCD)} detect particles directly from the tail of the
shower when it reaches the ground. These instruments have a very
large FoV, large duty-cycle but higher energy threshold and
usually have lower signal to noise ratios compared to IACTs~\citep{2015CRPhy..16..610D}.

@@

%% Context
Ground-based \gammaray astronomy has been historically conducted
by experiments operated by independent collaborations, each relying
on their own proprietary data and analysis software developed as part of the
instrument. While this model has been successful so far, it does not
permit easy combination of data from several instruments and therefore
limits the interoperability of existing facilities. This lack of
interoperability currently limits the full exploitation of the
available \gammaray data, especially because the different instruments often have
complementary sky coverages, and the various detection
techniques have complementary properties in terms of the energy range covered,
duty cycle and spatial resolution.

@@

%% \gammapy: concept and goals
%%% Question: starting date? Is is relevant, what is the real start?
The \gammapy project was started following the idea of \astropy with the objective of building a common
software library for very high-energy \gammaray data analysis \citep{gammapy_2015}. 
The core of the idea is illustrated in Figure~\ref{fig:big_picture}. Various \gammaray instruments
export their data to a standardised common data format -- the \gadf. This data can then be combined and
analysed using a single common software library.
This means that the \gammapy package is not specific to any instrument,
but an independent community developed software project.
The \gammapy package is is built on the scientific Python ecosystem: it uses \numpy~\citep{numpy} for n-dimensional data
structures, \scipy~\citep{2020SciPy-NMeth} for numerical algorithms, \astropy~\citep{astropy} for
astronomy-specific functionality, and \matplotlib~\citep{matplotlib} for visualization.

@@

%% CN: except for the figures we never introduced what the DL3 or the Data Level acronym is
%% maybe add (see Fig.~1)
\gammaray data at the Data Level 3 (DL3) therefore consist of lists of \gammaray-like events and their
corresponding instrument response functions. The latter include the effective area ($A_{\rm eff}$),
point spread function and energy dispersion ($E_{\rm disp}$).
In general, they depend on event's detector
geometrical parameters, e.g. the field-of-view location or the event elevation angle. So they might be parametrised as
function of such parameters specific to the instrumental technical.

@@

%% CN: I would actually distingusih two cases here:
%% 1) the Bkg is already provided as a parametrisation
%% 2) the Bkg is not provided and, thus, has to be estimated from the obs.
After all data has been projected into the same geometry, it is typically
required to improve the residual hadronic background estimate. As residual hadronic
background models can be subject to significant systematic uncertainties,
these models can be improved by taking into account actual data
from regions without known \gammaray sources. This includes methods 
such as the ring or the field-of-view background techniques or
background measurements performed within, e.g. reflected regions~\citep{Berge07}.
Data measured at the field-of-view or energy boundaries of the instrument are typically
associated with a systematic uncertainty in the IRF. For this reason this part 
of the data is often excluded from subsequent analysis by defining regions of
 \enquote{safe} data in the spatial as well as energy dimension.
All of these data reduction steps are performed by classes and functions
implemented in the \code{gammapy.makers} subpackage (see Section~\ref{ssec:gammapy-makers}).

@@

A typical usage example is shown in Figure~\ref{fig*:minted:gp_data}.
First a \code{DataStore} object is created from the path of the data
directory. The directory contains an observation as well as FITS HDU 
index file which assigns the correct data and IRF FITS files and HDUs
to the given observation ID. The \code{DataStore}
object gathers a collection of observations and provides ancillary
files containing information about the telescope observation mode and the
content of the data unit of each file. The \code{DataStore} allows for
selecting a list of observations based on specific filters.

@@

%% CN: ref. for King function https://ui.adsabs.harvard.edu/abs/2011A%26A...534A..34R/abstract
%% CN: I would not say that the PSFKing takes into account "the gamma and sigma", fo what?
%% I would put a formula or quickly describe the function (e.g. it is a power-law, an exponential...)
\subsubsection{Point Spread Function}
\gammapy allows user to treat different kinds of PSFs,
in particular, parametric multi-dimensional Gaussians (\code{EnergyDependentMultiGaussPSF})
or King profile functions (\code{PSFKing}). The \code{EnergyDependentMultiGaussPSF}
class is able to handle up to three
Gaussians, defined in terms of amplitudes and sigma given for each true energy
and offset angle bin. Similarly, \code{PSFKing} takes into account the gamma and
sigma parameters. The general \code{ParametricPSF} class allows users to create a
custom PSF with a parametric representation different from Gaussian(s) or King profile(s).
The generic \code{PSF3D} class stores a radial symmetric profile of a
PSF to represent non-parametric shapes, again depending on true energy
of offset form the pointing position.

@@

\subsubsection{Instrumental Background}
The instrumental background rate can be represented in \gammapy as either a 2-dimensional
data structure (\code{Background2D}) of count rate normalised per steradians and energy at different
reconstructed energies and offset angles or as rate per steradians and energy, as a
function of reconstructed energy and detector coordinates (\code{Background3D}).
In the former, the background is expected to follow a radially symmetric shape,
while in the latter, it can be more complex.

@@

\subsection{gammapy.makers}
\label{ssec:gammapy-makers}
%
\begin{figure}
	\small
    \import{code-examples/generated/}{gp_makers}
	\caption{
        Using \code{gammapy.makers} to reduce DL3 level data into a
		\code{MapDataset}. All \code{Maker} classes represent 
		a step in the data reduction process. They take
        the configuration on initialisation of the class. They 
		also consistently define \code{.run()} methods, which take
		a dataset object and optionally an \code{Observation} 
		object. In this way, \code{Maker} classes can be chained
		to define more complex data reduction pipelines. The output
		of the code example is shown in Figure~\ref{fig:code_example_gp_makers}.
    }
	\label{fig*:minted:gp_makers}
\end{figure}
%
The \code{gammapy.makers} sub-package contains the various classes and functions required
to process and prepare \gammaray data from the DL3 to the DL4, representing the input for modeling and fitting.
First, events are binned and IRFs are interpolated and projected onto the chosen analysis
geometry. The end product of the data reduction process are a set of binned counts,
background exposure, psf and energy dispersion maps at the DL4 level. 
The \code{MapDatasetMaker} and \code{SpectrumDatasetMaker} are
responsible for this task for 3D and 1D analyses, respectively (see Figure~\ref{fig*:minted:gp_makers}).

@@

To perform simple statistical estimations on counts measurements,
\code{CountsStatistic} classes encapsulate the aforementioned statistic functions to
measure excess counts and estimate the associated statistical significance,
errors and upper limits. They perform maximum likelihood ratio tests to
estimate significance (the square root of the statistic difference) and compute
likelihood profiles to measure errors and upper limits. The code example
\ref{codeexample:stats} shows how to compute the Li \& Ma
significance~\citep{LiMa} of a set of measurements.

@@

\subsubsection{Models}
\label{sssec:models}
Source models in \gammapy (Eq.~\ref{eq:general_source_model}) are four-dimensional 
analytical models which support two spatial dimensions defined by the sky coordinates
$\ell, b$, an energy dimension $E$, and a time dimension $t$. To simplify the the definition of the
models, \gammapy uses a factorised representation of the total source
model:

@@

The analytical spatial models are all normalized such as they integrate to
unity over the entire sky. The template spatial models may not, so in that special
case they have to be combined with a \code{NormSpectralModel}.

@@

In addition to the best-fit flux \textit{norm}, all estimators compute
quantities corresponding to this flux. This includes:
the predicted number of total, signal and background
counts per flux bin; the total fit statistics
of the best fit model; the fit statistics of the
null hypothesis; and the difference between both,
the so-called TS value.
From the TS value the significance of the measured signal and associated flux
can be derived.

@@

The sub-package also provides \texttt{matplotlib} implementations of specific
colormaps for false color image representation. Those colormaps have
been historically used by larger collaborations in the very high-energy
domain (such as \milagro or \hess) as \enquote{trademark} colormaps.
While we explicitly discourage the use of those colormaps for publication
of new results, because they do not follow modern visualization
standards, such as linear brightness gradients and accessibility
for visually impaired people, we still consider the colormaps
useful for reproducibility of past results.

@@

While the present list of use cases is rather preliminary, this can be enriched
with time with by users and/or developers according to future needs.

@@

The \magic data is included from the data level DL3. They consist of two
observations of $20\,{\rm min}$ each, chosen from the dataset used to estimate
the performance of the upgraded stereo system~\citep{magic_performance} and
already included in~\cite{joint_crab}. The observations were taken at small
zenith angles ($<30^{\circ}$) in wobble mode~\citep{fomin_1994}, with the
source sitting at an offset of $0.4^{\circ}$ from the FoV center. Their energy
range spans $80\,{\rm GeV} - 20\,{\rm TeV}$. They data reduction for the 1D analysis
is applied, and the data are reduced to a \code{SpectrumDataset} before being fitted.

@@

For the association and classification step, which is tightly connected to the
population studies, we can easily compare the fitted models to the set of
existing \gammaray catalogs available in \code{gammapy.catalog}. Further
multi-wavelength cross-matches are usually required to characterize the
sources. This can e.g. easily be achieved by relying on coordinate
handling from Astropy in combination with affiliated packages
\textit{Astroquery}~\citep{astroquery}.

Studies performed on simulations not only offer a first glimpse on what could
be the sky seen by CTA (according to our current knowledge on source
populations), but also give us the opportunity to test the software on complex
use cases\footnote{Note that the CTA-GPS simulations were performed with the
	\textit{ctools} package~\citep{2016A&A...593A...1K} and analysed with both
	\textit{ctools} and \textit{gammapy} packages in order to cross-validate
	them.}. So we can  improve performances, optimize our analyses strategies, and
identify the needs in term of parallelisation to process the large datasets
provided by the surveys.

>>>>

%% \gammaray astronomy
%% CN: I think we should use the expression Water Cherenkov Detectors and use the
%% abbreviation (WCD) all along the text, later we use "particle samplers"
The \gammaray range of the electromagnetic spectrum provides us insights into the
most energetic processes in the universe such as those accelerating particles in the surroundings of
black holes, and remnants of supernova explosions. As in other
branches of astronomy, \gammarays can be observed by
satellite as well as ground based instruments.
Ground-based instruments use the Earth's atmosphere as a particle detector.
Very-high-energy (VHE) cosmic \gammarays interact in the atmosphere and
create a large shower of secondary particles that can be observed from the ground.
Ground-based \gammaray astronomy relies on these extensive air showers to detect the
primary \gammaray photons and infer their incident direction and energy.
VHE \gammaray astronomy covers the energy range from few tens of ${\rm GeV}$ up to the ${\rm PeV}$.
There are two main categories of ground-based instruments:

@@

\textit{Water Cherenkov Detectors (WCD)} detect particles directly from the tail of the
shower when it reaches the ground. These instruments have a very
large FoV and large duty-cycle, but a higher energy threshold and
usually lower signal to noise ratios compared to IACTs~\citep{2015CRPhy..16..610D}.

@@

%% Context
Ground-based \gammaray astronomy has been historically conducted
through experiments operated by independent collaborations, each relying
on their own proprietary data and analysis software developed as part of the
instrument. While this model has been successful so far, it does not
permit easy combination of data from several instruments and therefore
limits the interoperability of existing facilities. This lack of
interoperability currently limits the full exploitation of the
available \gammaray data, especially because the different instruments often have
complementary sky coverages, and the various detection
techniques have complementary properties in terms of energy range covered,
duty cycle and spatial resolution.

@@

%% \gammapy: concept and goals
%%% Question: starting date? Is is relevant, what is the real start?
The \gammapy project was started following the idea of \astropy with the objective of building a common
software library for very high-energy \gammaray data analysis \citep{gammapy_2015}. 
The core of the idea is illustrated in Figure~\ref{fig:big_picture}. Various \gammaray instruments
export their data to the standardised common \gadf data format. This data can then be combined and
analysed using a single common software library.
This means that the \gammapy package is not specific to any instrument,
but an independent community developed software project.
The \gammapy package is built on the scientific Python ecosystem: it uses \numpy~\citep{numpy} for n-dimensional data
structures, \scipy~\citep{2020SciPy-NMeth} for numerical algorithms, \astropy~\citep{astropy} for
astronomy-specific functionality, \iminuit~\citep{iminuit} for numerical minimisation
and \matplotlib~\citep{matplotlib} for visualization.

@@

%% CN: except for the figures we never introduced what the DL3 or the Data Level acronym is
%% maybe add (see Fig.~1)
\gammaray data at the Data Level 3 (DL3) therefore consist of lists of \gammaray-like events and their
corresponding instrument response functions. The latter include the effective area ($A_{\rm eff}$),
point spread function and energy dispersion ($E_{\rm disp}$).
In general, IRFs depend on the geometrical parameters of the detector, e.g. location 
of an event in the FoV or the elevation angle of the event. So IRFs might be parametrised
as a function of such detector specific coordinates too.

@@

%% CN: I would actually distingusih two cases here:
%% 1) the Bkg is already provided as a parametrisation
%% 2) the Bkg is not provided and, thus, has to be estimated from the obs.
After all data have been projected into the same geometry, it is typically
required to improve the residual hadronic background estimate. As residual hadronic
background models can be subject to significant systematic uncertainties,
these models can be improved by taking into account actual data
from regions without known \gammaray sources. This includes methods 
such as the ring or the field-of-view background techniques or
background measurements performed within, e.g. reflected regions~\citep{Berge07}.
Data measured at the field-of-view or energy boundaries of the instrument are typically
associated with a systematic uncertainty in the IRF. For this reason this part 
of the data is often excluded from subsequent analysis by defining regions of
 \enquote{safe} data in the spatial as well as energy dimension.
All of these data reduction steps are performed by classes and functions
implemented in the \code{gammapy.makers} subpackage (see Section~\ref{ssec:gammapy-makers}).

@@

A typical usage example is shown in Figure~\ref{fig*:minted:gp_data}.
First a \code{DataStore} object is created from the path of the data
directory. The directory contains an observation as well as a FITS HDU 
index file which assigns the correct data and IRF FITS files and HDUs
to the given observation ID. The \code{DataStore}
object gathers a collection of observations and provides ancillary
files containing information about the telescope observation mode and the
content of the data unit of each file. The \code{DataStore} allows for
selecting a list of observations based on specific filters.

@@

%% CN: ref. for King function https://ui.adsabs.harvard.edu/abs/2011A%26A...534A..34R/abstract
%% CN: I would not say that the PSFKing takes into account "the gamma and sigma", fo what?
%% I would put a formula or quickly describe the function (e.g. it is a power-law, an exponential...)
\subsubsection{Point Spread Function}
\gammapy allows user to treat different kinds of PSFs,
in particular, parametric multi-dimensional Gaussians (\code{EnergyDependentMultiGaussPSF})
or King profile functions (\code{PSFKing}). The \code{EnergyDependentMultiGaussPSF}
class is able to handle up to three
Gaussians, defined in terms of amplitudes and sigma given for each true energy
and offset angle bin. Similarly, \code{PSFKing} takes into account the gamma and
sigma parameters. The general \code{ParametricPSF} class allows users to create a
custom PSF with a parametric representation different from Gaussian(s) or King profile(s).
The generic \code{PSF3D} class stores a radial symmetric profile of a
PSF to represent non-parametric shapes, again depending on true energy
and offset from the pointing position.

@@

\subsubsection{Instrumental Background}
The instrumental background rate can be represented as either a 2-dimensional
data structure named \code{Background2D} or a 3-dimensional one named \code{Background3D}.
The background rate is stored as a differential count rate, normalised per solid angle
and energy at different reconstructed energies. In the first case the background is
expected to follow a radially symmetric shape and changes only with the offset angle from
FoV center. In the other case the background is allowed to vary with 
longitude and latitude of a tangential FoV coordinates system.

@@

\subsection{gammapy.makers}
\label{ssec:gammapy-makers}
%
\begin{figure}
	\small
    \import{code-examples/generated/}{gp_makers}
	\caption{
        Using \code{gammapy.makers} to reduce DL3 level data into a
		\code{MapDataset}. All \code{Maker} classes represent 
		a step in the data reduction process. They take
        the configuration on initialisation of the class. They 
		also consistently define \code{.run()} methods, which take
		a dataset object and optionally an \code{Observation} 
		object. In this way, \code{Maker} classes can be chained
		to define more complex data reduction pipelines. The output
		of the code example is shown in Figure~\ref{fig:code_example_gp_makers}.
    }
	\label{fig*:minted:gp_makers}
\end{figure}
%
The \code{gammapy.makers} sub-package contains the various classes and functions required
to process and prepare \gammaray data from the DL3 to the DL4, representing the input for modeling and fitting.
First, events are binned and IRFs are interpolated and projected onto the chosen analysis
geometry. The end product of the data reduction process is a set of binned counts,
background exposure, psf and energy dispersion maps at the DL4 level. 
The \code{MapDatasetMaker} and \code{SpectrumDatasetMaker} are
responsible for this task for 3D and 1D analyses, respectively (see Figure~\ref{fig*:minted:gp_makers}).

@@

To perform simple statistical estimations on counts measurements,
\code{CountsStatistic} classes encapsulate the aforementioned statistic functions to
measure excess counts and estimate the associated statistical significance,
errors and upper limits. They perform maximum likelihood ratio tests to
estimate significance (the square root of the statistic difference) and compute
likelihood profiles to measure errors and upper limits. The code example
\ref{fig*:minted:gp_stats} shows how to compute the Li \& Ma
significance~\citep{LiMa} of a set of measurements.

@@

\subsubsection{Models}
\label{sssec:models}
Source models in \gammapy (Eq.~\ref{eq:general_source_model}) are four-dimensional 
analytical models which support two spatial dimensions defined by the sky coordinates
$\ell, b$, an energy dimension $E$, and a time dimension $t$. To simplify the definition of the
models, \gammapy uses a factorised representation of the total source
model:

@@

The analytical spatial models are all normalized such that they integrate to
unity over the entire sky. The template spatial models may not, so in that special
case they have to be combined with a \code{NormSpectralModel}.

@@

In addition to the best-fit flux \textit{norm}, all estimators compute
quantities corresponding to this flux. This includes:
the predicted number of total, signal and background
counts per flux bin; the total fit statistics
of the best fit model; the fit statistics of the
null hypothesis; and the difference between both,
the so-called \textit{TS value}.
From this \textit{TS value}, a significance of the measured signal and associated flux
can be derived.

@@

The sub-package also provides \texttt{matplotlib} implementations of specific
colormaps. Those colormaps have been historically used by larger collaborations
in the very high-energy domain (such as \milagro or \hess) as \enquote{trademark}
colormaps. While we explicitly discourage the use of those colormaps for publication
of new results, because they do not follow modern visualization
standards, such as linear brightness gradients and accessibility
for visually impaired people, we still consider the colormaps
useful for reproducibility of past results.

@@

While the present list of use cases is rather preliminary, this can be enriched
with time by users and/or developers according to future needs.

@@

The \magic data is included from the data level DL3. They consist of two
observations of $20\,{\rm min}$ each, chosen from the dataset used to estimate
the performance of the upgraded stereo system~\citep{magic_performance} and
already included in~\cite{joint_crab}. The observations were taken at small
zenith angles ($<30^{\circ}$) in wobble mode~\citep{fomin_1994}, with the
source sitting at an offset of $0.4^{\circ}$ from the FoV center. Their energy
range spans $80\,{\rm GeV} - 20\,{\rm TeV}$. The data reduction for the 1D analysis
is applied, and the data are reduced to a \code{SpectrumDataset} before being fitted.

@@

For the association and classification step, which is tightly connected to the
population studies, we can easily compare the fitted models to the set of
existing \gammaray catalogs available in \code{gammapy.catalog}. Further
multi-wavelength cross-matches are usually required to characterize the
sources. This can easily be achieved by relying on coordinate
handling from Astropy in combination with affiliated packages
\textit{Astroquery}~\citep{astroquery}.

Studies performed on simulations not only offer a first glimpse on what could
be the sky seen by CTA (according to our current knowledge on source
populations), but also give us the opportunity to test the software on complex
use cases\footnote{Note that the CTA-GPS simulations were performed with the
	\textit{ctools} package~\citep{2016A&A...593A...1K} and analysed with both
	\textit{ctools} and \textit{gammapy} packages in order to cross-validate
	them.}. So we can improve performance, optimize our analyse strategies, and
identify the needs in terms of parallelisation to process the large datasets
provided by surveys.
